A Match Made in Landfills?

### 2023-04-24 3 Analysis Scripts/Data_Exploratory_Analysis.qmd
Images, Manuscript Edits, etc
Take 1

#create correlation matrix
cor(variables, use = "complete.obs") #ignore NAs
corrplot(cor(variables, use = "complete.obs"))

@@

```{r}
stork_AMR_clean %>%
  ggplot(aes(as.factor(yr), sum))+
  geom_boxplot()
```

@@

```{r, message=FALSE, warning=FALSE}
stork_AMR_2021%>%
  ggplot(aes(mean_lui, nsuccess, color=as.factor(MDR)))+
  geom_point()+
  geom_smooth(method = "lm")
```

@@

```{r, warning=FALSE}
stork_AMR_2021 %>%
  ggplot(aes(as.factor(MDR), mean_lui))+
  geom_boxplot()
```

>>>>

#create correlation matrix
cor(variables, use = "complete.obs") #ignore NAs
CorrPlot <- corrplot(cor(variables, use = "complete.obs"))
CorrPlot

@@

```{r}
pSB <- stork_AMR_clean %>%
  ggplot(aes(as.factor(yr), sum))+
  geom_boxplot() +
  ylab('Breeding Season Sampling Period') +
  xlab ('Antimicrobial Gene Resistance Burden')

while (!is.null(dev.list()))  dev.off()
#Tiff 
tiff(file = here('5 Manuscript','manuscript','Images', 'Prelim_BurdenvsSamp.tiff'))
pSB
dev.off() 
```

@@

```{r, message=FALSE, warning=FALSE}
pLN <- stork_AMR_2021%>%
  ggplot(aes(mean_lui, nsuccess, color=as.factor(MDR)))+
  geom_point()+
  geom_smooth(method = "lm")+
  ylab('Nest Success (Includes 2020 to 2021)') +
  xlab ('Mean Landfill Use Index (LUI)')

#Tiff 
tiff(file = here('5 Manuscript','manuscript', 'Images', 'Prelim_LUIvsNS.tiff'))
pLN
dev.off() 
```

@@

```{r, warning=FALSE}
pLM <- stork_AMR_2021 %>%
  ggplot(aes(as.factor(MDR), mean_lui))+
  geom_boxplot()+
  ylab('Mean Landfill Use Index (LUI)') +
  xlab ('Multi-Drug Resistance')

#save as tiff
tiff(file = here('5 Manuscript','manuscript', 'Images', 'Prelim_LUIvsMD.tiff'))
pLM
dev.off() 
```

### 2023-04-24 3 Analysis Scripts/StatiscalAnalysis_ModelFitting.qmd
readme.md Files + Background
Missing bibTEX files (Currently troubleshooting)
Images, Manuscript Edits, etc
Take 1

MLp

ggsave(MLp, file="LUIvsMDR_LME.tiff")
```

@@

# save fit results table  
model2 %>% 
  saveRDS(here("4 Results","Models", "LMERmodel2.rds"))

@@

BLp

ggsave(BLp, file="LUIvsBurden_LME.tiff")
```

>>>>

##Not working... fix later
  while (!is.null(dev.list()))  dev.off()
#Tiff 
tiff(file = here('5 Manuscript','manuscript','Images', 'LUIvsMDR_LME.tiff'))
MLp
dev.off() 
```

@@

# save fit results table  
model2 %>% 
  saveRDS(here("4 Results/Models/LMERmodel2.rds"))

@@

while (!is.null(dev.list()))  dev.off()
#Tiff 
tiff(file = here('5 Manuscript','manuscript','Images', 'LUIvsBurden_LME.tiff'))
BLp
dev.off() 
```

### 2023-04-24 5 Manuscript/manuscript/Manuscript.qmd
Images, Manuscript Edits, etc
Take 1

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# load a few R packages
library(here)
library(knitr)
library(knitcitations)
```

@@

## General Background Information (WILL FIGHT WITH BIBTEX TOMORROW)

@@

Birds have a long history of being used as indicators for environmental and human health. For example, wading bird surveillance is commonplace in south Florida to monitor heavy metal levels in the greater everglades ecosystem. Birds are also often utilized in zoonotic pathogen surveillance (i.e., USDA sentinel chickens to monitor West Nile Virus). In recent years, a similar approach has been utilized to study AMR prevalence/impacts on different landscapes. Currently, the carriage of AMR genes presents an unknown risk to the individual birds and may have greater implications for free-living avian communities and conservation. Resistance genes are thought to alter the fitness of a pathogen [@Friedman2016], and have been correlated with increased virulence [@Escudeiro2019], in turn increasing morbidity and mortality in affected hosts. Finally, at the population management level, birds have also been held ‘responsible’ for the dissemination of enteric pathogens (e.g., *Escherichia coli*, *Salmonella* spp.) that have caused outbreaks in produce, yet, they are only moving these bacteria from areas already contaminated by human activities [@Hoelzer2011]. Thus, understanding how avian behaviors that overlap with risky environmental factors (e.g., foraging in agricultural areas or landfills) impact the carriage of resistance, is key because: 1) it allows us to better understand the role birds play in AMR dissemination, and  2) it informs targeted population management and conservation strategies—such as discouraging birds from utilizing certain areas.

In 2019, the USGS revealed that more than half of GPS-tracked gulls (*Larus* ssp.) had acquired antimicrobial resistant *E. coli* from landfills and disseminated it long distances to pristine, unaltered habitats [@Ahlstrom2019]. This supports the belief that birds who utilize and move between human-dominated and natural landscapes may become both reservoirs and efficient transporters of resistance determinants to natural environments.For wildlife in general, resistance prevalence tends to increase with proximity to human populations [@Niu2013]. It is believed that animals that rely on anthropogenic resources are often concentrated at high densities that promote the large-scale mixing of bacteria, encouranging horizontal gene transfer (HGT) facilitating the rise of novel resistant strains [@Wellington2013]. For instance, birds that encounter human waste (e.g., by drinking contaminated water) can acquire resistance genes through the exchange of naturally present genes in the bacteria they carry, with the genes in the waste—this can co-select for mobile genetic elements carrying multiple resistant genes [@Wellington2013]. Moreover, ecological factors such as migration and high densities during breeding season may increase acquisition rates and the dissemination of AMR . The western white stork are one such species that heavily utilizes agricultural and urbam areas, providing a unique opportunity to explore how anthropogenic land use influences the carriage of antimicrobial resistance genes.

@@

## Questions/Hypotheses to be addressed

Does landfill use by storks increase the likelihood of carrying Antimicrobial resistance (AMR) genes and resistance gene burden?.

Stork that visit the landfill more often will have a higher likelihood of carrying AMR genes and have a higher burden of AMR genes.

Is nest success related to AMR gene burden and multi-drug resistance (i.e., resistance to 3 or more drug classes)?

Stork with more AMR gene burdens and multi-drug resistance will have lower nest success.

@@

## Data import and cleaning (Reference file and give details HERE. Reference a ReadME/doc here)

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.*

```{r, warning=FALSE}
library(tidyverse)
library(here)

@@

#### (Note: still trying to figure out how to do this with the performance package) We evaluated model fit for all models using Akaike Information Criterion scores adjusted for small sample sizes (AICc: Burnham and Anderson, 2002). The best fit model was selected by having a ΔAICc \< 2.\_

@@

## Exploratory/Descriptive analysis (SUMMARY STATS IN EXPLORATION FILE TBD -\> Burden/MDR associated with each class. summary table? with actual summary stats)

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@@

```{r}
#| label: fig-result
#| fig-cap: "Linear Mixed Model Output with 95% CI."
#| echo: FALSE
knitr::include_graphics(here("3 Analysis Scripts","LUIvsBurden_LME.tiff"))

#| label: Figure2
#| fig-cap: "Multi-drug resistance as predicted by the variables landfill use index and age."
#| echo: FALSE
knitr::include_graphics(here("3 Analysis Scripts","LUIvsMDR_LME.tiff"))
```

@@

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

@@

@@

@@

*What are the main take-home messages?*

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# load a few R packages
library(here)
library(knitr)
library(knitcitations)
library(tidyverse)
```

@@

## General Background Information

@@

Birds have a long history of being used as indicators for environmental and human health. For example, wading bird surveillance is commonplace in south Florida to monitor heavy metal levels in the greater everglades ecosystem. Birds are also often utilized in zoonotic pathogen surveillance (i.e., USDA sentinel chickens to monitor West Nile Virus). In recent years, a similar approach has been utilized to study AMR prevalence/impacts on different landscapes. Currently, the carriage of AMR genes presents an unknown risk to the individual birds and may have greater implications for free-living avian communities and conservation. Resistance genes are thought to alter the fitness of a pathogen [@Friedman2016], and have been correlated with increased virulence [@Escudeiro2019], in turn increasing morbidity and mortality in affected hosts. Finally, at the population management level, birds have also been held 'responsible' for the dissemination of enteric pathogens (e.g., *Escherichia coli*, *Salmonella* spp.) that have caused outbreaks in produce, yet, they are only moving these bacteria from areas already contaminated by human activities [@Hoelzer2011]. Thus, understanding how avian behaviors that overlap with risky environmental factors (e.g., foraging in agricultural areas or landfills) impact the carriage of resistance, is key because: 1) it allows us to better understand the role birds play in AMR dissemination, and 2) it informs targeted population management and conservation strategies---such as discouraging birds from utilizing certain areas.

In 2019, the USGS revealed that more than half of GPS-tracked gulls (*Larus* ssp.) had acquired antimicrobial resistant *E. coli* from landfills and disseminated it long distances to pristine, unaltered habitats [@Ahlstrom2019]. This supports the belief that birds who utilize and move between human-dominated and natural landscapes may become both reservoirs and efficient transporters of resistance determinants to natural environments.For wildlife in general, resistance prevalence tends to increase with proximity to human populations [@Niu2013]. It is believed that animals that rely on anthropogenic resources are often concentrated at high densities that promote the large-scale mixing of bacteria, encouranging horizontal gene transfer (HGT) facilitating the rise of novel resistant strains [@Wellington2013]. For instance, birds that encounter human waste (e.g., by drinking contaminated water) can acquire resistance genes through the exchange of naturally present genes in the bacteria they carry, with the genes in the waste---this can co-select for mobile genetic elements carrying multiple resistant genes [@Wellington2013]. Moreover, ecological factors such as migration and high densities during breeding season may increase acquisition rates and the dissemination of AMR . The western white stork are one such species that heavily utilizes agricultural and urbam areas, providing a unique opportunity to explore how anthropogenic land use influences the carriage of antimicrobial resistance genes.

@@

## Objective

Evaluate both antimicrobial gene resistence diversity and burden in white storks that utilize landfills and natural areas over the breeding season of years 2020 to 2021.

## Hypotheses

Increased landfill use will increase both the diversity and burden of antimicrobial resistance genes in white stork.

Nest success will be negatively affected by antimicrobial gene burden and multi-drug resistance (i.e., resistance to 3 or more drug classes).

@@

## Data import and cleaning

Raw data was collected during both field seasons (2020 and 2021) and maintained on an excel file. This excel file was amended in 2021 to include the laboratory results from the AMR qPCR. The raw excel file can be found in the "1 Data Cleaning Script" folder in this [projects repository](https://github.com/rfranci3/SETHLATTERRAQUELFRANCISCO-MADA-project). Cleaned data was evaluated for normality and each variable was standardized and checked correlative relationships. All variables were kept as they did not appear strongly correlated.

@@

@@

## Exploratory/Descriptive analysis

Cleaned data was then visually explored to evaluate trends. Notably, it did not seem to appear that multi-drug resistance played a factor in nest success (Figure 2). However, sampling periods did appear important to the amount of ARG burden found in white stork feces (Figure 3). The most notable trend in the data however, was the appearance that multi-drug resistence in fact declined with the increase of landfill use by the adult white storks (Figure 4).

![Figure 2: Multi-drug resistance and nest success (i.e., number of chicks fledged) for each nest during the 2020 and 2021 white stork breeding seasons.](Images/Prelim_LUIvsNS.tiff)

![Figure 3: Antimicrobial Drug Resistence burden across all three sampling periods during the 2020 and 2021 white stork breeding seasons.](Images/Prelim_BurdenvsSamp.tiff)

![Figure 4: Multi-drug resistance and Landfill Use Index for each nest during the 2020 and 2021 white stork breeding seasons.](Images/Prelim_LUIvsMD.tiff)

@@

@@

@@

![Figure 5: Multi-drug resistance as predicted by the variables landfill use index and age during the 2020 and 2021 white stork breeding seasons. The gray is a 95% CI](Images/LUIvsMDR_LME.tiff)

@@

![Figure 6: Anti-microbial resistance gene burden of each nest as predicted by the variables landfill use index and sampling period during the 2020 and 2021 white stork breeding seasons. The gray is a 95% CI](Images/LUIvsBurden_LME.tiff)

@@
